---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Jamieson Nathan"
date: "01/21/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}


library(tidyverse) 
library(dsbox)
library(mosaicData) 
library(ggplot2)

staff <- read_csv("data/instructional-staff.csv")

```

### 1. Include the line plot you made above in your report and make sure the figure width is large enough to make it legible (also fix the title, axis labels, and legend label)

```{r modification1}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

```{r graph, fig.width = 8}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()+
  theme_bw()+
  labs(
    x = "Year",
    y = "Percent of Total Instructional Staff",
    title = "Trends in Instructional Staff Employee Status",
    color = "Faculty Status"
  )
```

### 2. Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types.

```{r revised-plot}

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year", values_to = "value") %>%
  mutate(
    value = as.numeric(value)
  )

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type  
  )) +
  geom_line(size = 1) +  
  scale_color_manual(  
    values = c(
      "Part-Time Faculty" = "firebrick",
      "Full-Time Tenured Faculty" = "navy",
      "Full-Time Tenure-Track Faculty" = "navy",
      "Full-Time Non-Tenure-Track Faculty" = "navy",
      "Graduate Student Employees" = "darkgrey"
    )
  ) +
  theme_classic() +  
  labs(
    title = "Trend Analysis of Instructional Staff Types Over Time",
    x = "Academic Year",
    y = "Percentage of Instructional Staff",
    color = "Faculty Type"  
  ) +
  geom_text(aes(label = ifelse(year == "2019", faculty_type, NA)), hjust = 1.1, vjust = 0) 


```

I adjusted the graph to improve clarity and interpretation. By linking colors to faculty types and using a legend, it's easier to differentiate the data. Custom colors emphasize trends, particularly the increase in part-time faculty. Labels on the latest data point highlight recent trends without cluttering the graph, ensuring the visualization is informative and visually appealing.


### 3. Fisheries

These visuals are very confusing and not super informative... 


```{r preparing-fishies}

library(ggplot2)
library(readr)
library(dplyr)  

fisheries <- read_csv("data/fisheries.csv")

str(fisheries)

head(fisheries)


```
```{r filter_data}

top_capture <- fisheries %>%
  arrange(desc(capture)) %>%
  slice(1:10)

top_aquaculture <- fisheries %>%
  arrange(desc(aquaculture)) %>%
  slice(1:10)

```

```{r visuals-fishies}

ggplot(top_capture, aes(x = reorder(country, capture), y = capture, fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Countries by Capture Production in 2016",
    x = "Country",
    y = "Production (Tons)",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

ggplot(top_aquaculture, aes(x = reorder(country, aquaculture), y = aquaculture, fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Countries by Aquaculture Production in 2016",
    x = "Country",
    y = "Production (Tons)",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


```

# Smokers in Whickham

### What type of study do you think these data come from: observational or experiment? Why?

### How many observations are in this dataset? What does each observation represent?

### How many variables are in this dataset? What type of variable is each? Display each variable using an appropriate visualization.

### What would you expect the relationship between smoking status and health outcome to be?


```{r initial-questions}

library(tidyverse)

data(Whickham)

?Whickham

```

Likely observational, as it would be pretty unethical to have an experimental control forcing people to smoke...

There are 1314 observation refering to different female participants in the study. There are 3 variables including outcome (survival sttus after 20 years; alive/dead), smoking status at baseline (yes/no), and age (in years at onset). 

Just based on common-sense, it would be likely that smokers at baseline die more than their counterparts at the 20-year follow-up. Although depending on age of initial testing this may not be as clear (i.e., 40-year old smokers are probably still doing okay as opposed to 60+). 

### Create a visualization depicting the relationship between smoking status and health outcome. Briefly describe the relationship, and evaluate whether this meets your expectations. Additionally, calculate the relevant conditional probabilities to help your narrative.


```{r smoker-survival-rates}

Whickham %>%
  count(smoker, outcome)

Whickham %>%
  count(smoker, outcome) %>%
  ggplot(aes(x = outcome, y = n, group = smoker, color = smoker)) +
  geom_line(linewidth = 2) +
  labs(
    title = "Survival Rates based on Smoking Status",
    x = "",
    y = "Number of participants",
    color = "Smoking Status"
  ) +
  theme_minimal()

Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(cond_prob = n /sum(n))

```

Contrary to expectations, these lines seem somewhat unrelated, and the conditional probability supports this. There does not seem to be much of a relationship between smoking status and survival rates at a 20-year followup. Perhaps Big Tobacco was right and cig's are good for you... Or we need a longer follow-up. 

### Re-create the visualization depicting the relationship between smoking status and health outcome, faceted by age_cat. What changed? What might explain this change? Extend the contingency table from earlier by breaking it down by age category and use it to help your narrative. We can use the contingency table to examine how the relationship between smoking status and health outcome differs between different age groups. This extension will help us better understand the patterns we see in the visualization, and explain any changes we observe.

```{r smokers-and-age}

Whickham_ages <- Whickham %>%
 mutate(age_cat = case_when(
    age <= "44" ~ "18-44",
    age > "44" & age <= "64" ~ "45-64",
    age > "64" ~ "65+",
      ))

Whickham_ages %>%
 ggplot(aes(
   y = fct_rev(smoker), 
   fill = fct_rev(outcome))) +
  geom_bar(position = "fill") + 
   facet_wrap(~age_cat)+
  labs(title = "Survival Rates",
     y = "Smoking Status", x = NULL)


```



```{r onto-the-next}

Whickham_ages %>%
  count(smoker, age_cat, outcome) %>%
  group_by(smoker, age_cat) %>%
  mutate(cond_prob = n /sum(n))

```

The plot and table show a more nuanced relationship than before, with older adults being less likely to survive at follow-up. However, the differences do seem small, maybe a line plot would have made more sense than a bar chart... All-in-all what I am taking away is that smoking is good for you, thanks Mason!
